---
title: "Fishing Activity in the Antarctic Peninsula"
subtitle: "State of the Antarctic Environment report"
author: " Pablo Deschepper (biodiversity.aq) - Institute for Natural Sciences, Belgium"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    embed-resources: true
    fig-width: 10
    fig-height: 7
    fig-dpi: 300
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
  cache: false   # we manage caching manually via saveRDS/readRDS
---

```{r, output = TRUE, eval = TRUE, echo = FALSE}
#| label: setup

# ── Package loading ────────────────────────────────────────────────────────────
# Install once if needed:
# install.packages(c("gfwr", "tidyverse", "sf", "scico",
#                    "patchwork", "ggrepel", "rnaturalearth",
#                    "rnaturalearthdata", "glue"))

library(gfwr)
library(tidyverse)
library(sf)
library(scico)
library(patchwork)
library(ggrepel)
library(rnaturalearth)
library(rnaturalearthdata)
library(glue)
library(stringr)

# ── Authentication ─────────────────────────────────────────────────────────────
# Add to ~/.Renviron:  GFW_TOKEN=your_token_here
# Then restart R. Obtain a free token at https://globalfishingwatch.org/our-apis/
key <- gfw_auth()

# ── Region: CCAMLR convention area ────────────────────────────────────────────

# Look up the numeric CCAMLR region id:
ccamlr_id <- gfw_region_id(region = "CCAMLR", region_source = "RFMO")
message("CCAMLR region id: ", ccamlr_id$id[1])
CCAMLR_ID <- ccamlr_id$id[1]   # numeric id used in all API calls

# ── Time-window parameters ────────────────────────────────────────────────────
CURRENT_YEAR  <- 2025
HEATMAP_YEARS <- (CURRENT_YEAR - 9):CURRENT_YEAR   # 5 most recent full years
TREND_START   <- HEATMAP_YEARS[1]

# ── gfwr column name constants ────────────────────────────────────────────────
# The API always returns these exact names (with spaces):
COL_LAT   <- "Lat"
COL_LON   <- "Lon"
COL_TIME  <- "Time Range"
COL_HOURS <- "Apparent Fishing Hours"

# ── Shared plot theme ──────────────────────────────────────────────────────────
theme_antarctic <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.title       = element_text(face = "bold", size = 14),
      plot.subtitle    = element_text(colour = "grey40", size = 11),
      plot.caption     = element_text(colour = "grey55", size = 8,
                                      hjust = 0, margin = margin(t = 8)),
      legend.position  = "right",
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey90", linewidth = 0.3),
      strip.text       = element_text(face = "bold", size = 11)
    )
}

GFW_CAPTION <- paste0(
  "Data source: Global Fishing Watch (globalfishingwatch.org) — ",
  "AIS-based apparent fishing effort (public-global-fishing-effort:v3.0), ",
  "CCAMLR convention area. Attribute: Global Fishing Watch."
)
```

# Introduction

This chapter presents spatiotemporal data products on commercial fishing
activity in the Antarctic Peninsula. Data are sourced from the **Global Fishing Watch (GFW)** open API (v3), which detects *apparent fishing effort* by analysing AIS vessel
broadcasts and classifying positions as fishing or non-fishing based on
changes in speed and heading. Effort is expressed as **fishing hours** per
0.1° × 0.1° grid cell.

::: {.callout-note}
**Data caveats:** AIS coverage approaches ~90% for vessels >24 m but is
substantially lower for smaller vessels. Results represent a lower bound
on total fishing activity.
:::

---

# Annual Fishing Hotspot Heatmaps (last `r length(HEATMAP_YEARS)` years)



```{r, output = FALSE, eval = TRUE, echo = FALSE}
#| label: fetch-heatmap-data

## Data retrieval

# ── Robust single-year fetch with get_last_report() fallback ──────────────────
#
# Large regional queries sometimes trigger an HTTP 524 timeout on the first
# call. When this happens, the API continues processing the report server-side.
# gfw_get_last_report() polls for and retrieves that completed report.
# We retry up to MAX_RETRIES times with a WAIT_SEC pause between attempts.

fetch_year_robust <- function(year, region_id, key,
                              group       = "GEARTYPE",
                              res         = "YEARLY",
                              max_retries = 5,
                              wait_sec    = 90) {

  message(glue("  Fetching {year} [{group} / {res}]..."))

  result <- tryCatch(
    gfw_ais_fishing_hours(
      spatial_resolution  = "LOW",
      temporal_resolution = res,
      group_by            = group,
      start_date          = glue("{year}-01-01"),
      end_date            = glue("{year}-12-31"),
      region              = region_id,
      region_source       = "RFMO",
      key                 = key
    ),
    error = function(e) {
      message(glue("    Initial call error: {conditionMessage(e)}"))
      NULL
    }
  )

  # If the first call failed (timeout / 429), poll gfw_get_last_report()
  if (is.null(result) || nrow(result) == 0) {
    message(glue("    Initial call returned no data. ",
                 "Polling gfw_get_last_report() (max {max_retries} tries)..."))

    for (attempt in seq_len(max_retries)) {
      message(glue("    Waiting {wait_sec} s before attempt {attempt}..."))
      Sys.sleep(wait_sec)

      result <- tryCatch(
        gfw_get_last_report(key = key),
        error = function(e) {
          message(glue("    Attempt {attempt} failed: {conditionMessage(e)}"))
          NULL
        }
      )

      # Accept result only if it has actual data rows
      if (!is.null(result) && is.data.frame(result) && nrow(result) > 0) {
        message(glue("    Success on attempt {attempt}: {nrow(result)} rows."))
        break
      }
    }
  }

  # Final guard: return NULL rather than an empty frame so compact() drops it
  if (is.null(result) || !is.data.frame(result) || nrow(result) == 0) {
    warning(glue("Could not retrieve data for {year}. Skipping."))
    return(NULL)
  }

  result
}

# ── Disk cache ─────────────────────────────────────────────────────────────────
# Cache stores the raw combined data frame. Delete the file to force a refresh.
cache_heatmap <- "cache_heatmap_data.rds"

if (file.exists(cache_heatmap)) {
  existing <- readRDS(cache_heatmap)

  # VALIDATE the cache — reject if empty (was written from a failed fetch)
  if (nrow(existing) == 0 || ncol(existing) == 0) {
    message("Cache is empty or corrupt — deleting and re-fetching.")
    file.remove(cache_heatmap)
  } else {
    message(glue("Valid cache found ({nrow(existing)} rows). Loading."))
    heatmap_raw <- existing
  }
}

if (!exists("heatmap_raw")) {
  message("Fetching heatmap data from GFW API (one call per year)...")

  raw_list <- map(
    HEATMAP_YEARS,
    ~ fetch_year_robust(as.vector(.x), region_id = CCAMLR_ID, key = key)
  )

  heatmap_raw <- map2(raw_list, HEATMAP_YEARS, function(df, yr) {
    if (is.null(df)) return(NULL)
    df |> mutate(year = as.integer(yr))
  }) |>
    compact() |>
    bind_rows()

  if (nrow(heatmap_raw) == 0) stop("All year fetches failed. Check your API token and connection.")

  saveRDS(heatmap_raw, cache_heatmap)
  message(glue("Heatmap data cached: {nrow(heatmap_raw)} rows -> {cache_heatmap}"))
}

glimpse(heatmap_raw)
```



```{r, output = FALSE, eval = TRUE, echo = FALSE}
#| label: wrangle-heatmap

## Data wrangling

# Rename space-containing column names to clean snake_case equivalents
heatmap_clean <- heatmap_raw |>
  rename(
    lat           = all_of(COL_LAT),
    lon           = all_of(COL_LON),
    time_range    = all_of(COL_TIME),
    fishing_hours = all_of(COL_HOURS)
  )

# Aggregate across gear types: total fishing hours per grid cell x year
heatmap_total <- heatmap_clean |>
  group_by(year, lat, lon) |>
  summarise(fishing_hours = sum(fishing_hours, na.rm = TRUE),
            .groups = "drop") |>
  # Log10(x+1) transform: right-skewed effort data; +1 avoids log(0)
  mutate(log_hours = log10(fishing_hours + 1))

# Gear-type breakdown for the composition section
gear_annual <- heatmap_clean |>
  rename(geartype = any_of("geartype")) |>
  group_by(year, geartype) |>
  summarise(fishing_hours = sum(fishing_hours, na.rm = TRUE),
            .groups = "drop") |>
  arrange(year, desc(fishing_hours))

message("Years in heatmap data: ",
        paste(sort(unique(heatmap_total$year)), collapse = ", "))
```

## Heatmap visualisation

```{r, output = TRUE, eval = TRUE, echo = FALSE}
#| label: fig-heatmaps
#| fig-cap: "Annual fishing effort hotspots in the CCAMLR convention area (Antarctic Peninsula), last `r length(HEATMAP_YEARS)` years. Colour encodes apparent fishing hours per 0.1° grid cell on a log₁₀ scale; all panels share the same colour axis for direct comparison."
#| fig-height: 16

world_sf <- ne_countries(scale = "medium", returnclass = "sf")

so_coord <- coord_sf(
  xlim = c(-80, -35 ), ylim = c(-75, -55),
  expand = FALSE, crs = st_crs(4326)
)

log_max <- max(heatmap_total$log_hours, na.rm = TRUE)

make_heatmap_panel <- function(yr) {
  df_yr <- filter(heatmap_total, year == yr)

  ggplot() +
    geom_tile(data = df_yr,
              aes(x = lon, y = lat, fill = log_hours),
              linewidth = 0) +
    geom_sf(data = world_sf,
            fill = "grey25", colour = "grey15", linewidth = 0.15) +
    scico::scale_fill_scico(
      palette  = "lajolla",
      na.value = "transparent",
      name     = "log₁₀(hrs+1)",
      limits   = c(0, log_max),
      guide    = guide_colorbar(title.position = "top",
                                barwidth  = unit(0.35, "cm"),
                                barheight = unit(3.5, "cm"))
    ) +
    so_coord +
    labs(title = as.character(yr)) +
    theme_antarctic() +
    theme(axis.title  = element_blank(),
          axis.text   = element_text(size = 7),
          plot.title  = element_text(size = 12, face = "bold", hjust = 0.5),
          legend.title = element_text(size = 8),
          legend.text  = element_text(size = 7))
}

panels <- map(HEATMAP_YEARS, make_heatmap_panel)

spatial_panels <- wrap_plots(panels, ncol = 2) +
  plot_annotation(
    title    = "Annual Fishing Effort Hotspots — Antarctic Peninsula (CCAMLR)",
    subtitle = glue("AIS-based apparent fishing hours per 0.1° grid cell, ",
                    "{min(HEATMAP_YEARS)}–{max(HEATMAP_YEARS)}"),
    caption  = GFW_CAPTION,
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(colour = "grey40", size = 12),
      plot.caption  = element_text(colour = "grey55", size = 8, hjust = 0)
    )
  )

plot(spatial_panels)
```

---

# Temporal Evolution of Fishing Intensity from `r TREND_START` to `r CURRENT_YEAR`


```{r, output = FALSE, eval = TRUE, echo = FALSE}
#| label: fetch-trend-data

## Data retrieval

cache_trend <- "cache_trend_data.rds"

if (file.exists(cache_trend)) {
  existing_t <- readRDS(cache_trend)
  if (nrow(existing_t) == 0 || ncol(existing_t) == 0) {
    message("Trend cache is empty or corrupt — deleting and re-fetching.")
    file.remove(cache_trend)
  } else {
    message(glue("Valid trend cache found ({nrow(existing_t)} rows). Loading."))
    trend_raw <- existing_t
  }
}

if (!exists("trend_raw")) {
  message("Fetching trend data from GFW API (one call per year)...")

  raw_trend_list <- map(
    TREND_START:CURRENT_YEAR,
    ~ fetch_year_robust(.x,
                        region_id = CCAMLR_ID,
                        key       = key,
                        group     = "FLAG",
                        res       = "MONTHLY")
  )

  trend_raw <- map2(raw_trend_list, TREND_START:CURRENT_YEAR, function(df, yr) {
    if (is.null(df)) return(NULL)
    df |> mutate(year = as.integer(yr))
  }) |>
    compact() |>
    bind_rows()

  if (nrow(trend_raw) == 0) stop("All trend fetches failed. Check your API token.")

  saveRDS(trend_raw, cache_trend)
  message(glue("Trend data cached: {nrow(trend_raw)} rows -> {cache_trend}"))
}

glimpse(trend_raw)
```


```{r, output = TRUE, eval = TRUE, echo = FALSE}
#| label: wrangle-trend

## Data wrangling

trend_clean <- trend_raw |>
  rename(
    lat           = all_of(COL_LAT),
    lon           = all_of(COL_LON),
    time_range    = all_of(COL_TIME),
    fishing_hours = all_of(COL_HOURS)
  )

# Filter for Antarctic Peninsula

trend_clean <- trend_clean %>%
  filter(lat > -80, lat < -35,
         lon > -75, lon < -55)

# Parse Time Range into Date
trend_dated <- trend_clean |>
  mutate(
    cal_year  = str_replace(time_range, "^(\\d{4})-\\d{2}$", "\\1"),
    cal_month  = as.factor(str_replace(time_range, "^\\d{4}-(\\d{2})$", "\\1")),
    month_lbl = factor(month.abb[cal_month], levels = month.abb)
  )

# Monthly totals: sum across all flags and grid cells
monthly_totals <- trend_dated |>
  group_by(cal_year, cal_month, month_lbl) |>
  summarise(fishing_hours = sum(fishing_hours, na.rm = TRUE),
            .groups = "drop") 

# Annual totals
annual_totals <- monthly_totals |>
  group_by(cal_year) |>
  summarise(annual_hours = sum(fishing_hours), .groups = "drop")

# Standardised monthly anomaly (z-score vs long-term monthly climatology)
monthly_clim <- monthly_totals |>
  group_by(cal_month) |>
  summarise(clim_mean = mean(fishing_hours, na.rm = TRUE),
            clim_sd   = sd(fishing_hours,   na.rm = TRUE),
            .groups   = "drop")

monthly_anom <- monthly_totals |>
  left_join(monthly_clim, by = "cal_month") |>
  mutate(anomaly_std = if_else(clim_sd > 0,
                               (fishing_hours - clim_mean) / clim_sd,
                               0))

message("Trend range: ", min(monthly_totals$date),
        " to ", max(monthly_totals$date))
```


```{r, output = FALSE, eval = TRUE, echo = FALSE}
#| label: fig-annual-trend
#| fig-cap: "Total apparent fishing hours in the CCAMLR convention area per year. Blue ribbon = LOESS smooth (span = 0.9) with 95% CI."

## Annual total fishing hours

annual_totals <- annual_totals |>
  mutate(cal_year = as.integer(cal_year))

p_annual <- ggplot(annual_totals,
                   aes(x = cal_year, y = annual_hours / 1e3)) +
  geom_smooth(method = "loess", span = 0.9,
              colour = "#1B6CA8", fill = "#1B6CA8",
              alpha = 0.15, linewidth = 0.9) +
  geom_line(colour = "grey55", linewidth = 0.6, linetype = "dashed") +
  geom_point(aes(colour = annual_hours / 1e6), size = 4.5,
             show.legend = FALSE) +
  geom_text_repel(
    aes(label = paste0(round(annual_hours / 1e3, 1), "")),
    size = 3, colour = "grey30",
    nudge_y = 0.06 * max(annual_totals$annual_hours / 1e3),
    min.segment.length = 0.4
  ) +
  scico::scale_colour_scico(palette = "bamako", direction = -1) +
  scale_x_continuous(breaks = TREND_START:CURRENT_YEAR) +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA),
                     expand = expansion(mult = c(0, 0.12))) +
  labs(
    title    = "Total Annual Fishing Effort — Antarctic Peninsula (CCAMLR)",
    subtitle = glue("Apparent fishing hours (all AIS-equipped fishing vessels), ",
                    "{TREND_START}–{CURRENT_YEAR}"),
    x = NULL, y = "Fishing hours (thousands)",
    caption = GFW_CAPTION
  ) +
  theme_antarctic()

p_annual
```


```{r, output = FALSE, eval = TRUE, echo = FALSE}
#| label: fig-monthly-anomaly
#| fig-cap: "Standardised monthly fishing effort anomaly (z-score vs long-term monthly mean). Blue = below average; red = above average. Numbers show rounded z-scores."

## Monthly anomaly heatmap

p_anom <- monthly_anom |>
  ggplot(aes(x = factor(cal_year), y = month_lbl, fill = anomaly_std)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%+.1f", anomaly_std)),
            size = 2.8, colour = "grey15") +
  scico::scale_fill_scico(
    palette = "vik",
    limits  = c(-3, 3),
    oob     = scales::squish,
    name    = "Z-score",
    guide   = guide_colorbar(title.position = "top",
                             barwidth  = unit(7, "cm"),
                             barheight = unit(0.5, "cm"))
  ) +
  scale_y_discrete(limits = rev(month.abb)) +
  labs(
    title    = "Monthly Fishing Effort Anomaly — Antarctic Peninsula (CCAMLR)",
    subtitle = glue("Standardised deviation from long-term monthly mean, ",
                    "{TREND_START}–{CURRENT_YEAR}"),
    x = NULL, y = NULL,
    caption = GFW_CAPTION
  ) +
  theme_antarctic() +
  theme(legend.position = "bottom",
        panel.grid      = element_blank(),
        axis.text.x     = element_text(angle = 45, hjust = 1))

p_anom
```

# Annual and monthly trends in fisheries intensities

```{r, output = TRUE, eval = TRUE, echo = FALSE}
#| label: fig-combined
#| fig-cap: "(A) Annual fishing effort totals and (B) month-by-year anomaly heatmap."
#| fig-height: 13

temporal_compo <- p_annual / p_anom +
  plot_annotation(tag_levels = "A",
                  theme = theme(plot.tag = element_text(face = "bold", size = 14)))

plot(temporal_compo)
```

---

# Gear-type composition over time

```{r, output = TRUE, eval = TRUE, echo = FALSE}
#| label: fig-gear-trend
#| fig-cap: "Annual fishing effort by gear type in the CCAMLR convention area."

gear_plot <- gear_annual |>
  mutate(
    gear_clean = case_when(
      geartype %in% c("trawlers", "drifting_longlines", "set_longlines",
                       "squid_jigger", "pole_and_line", "purse_seines",
                       "set_gillnets", "pots_and_traps") ~ geartype,
      TRUE ~ "other"
    ),
    gear_clean = str_replace_all(gear_clean, "_", " ") |> str_to_title()
  ) |>
  group_by(year, gear_clean) |>
  summarise(fishing_hours = sum(fishing_hours, na.rm = TRUE), .groups = "drop")

gearplot <- ggplot(gear_plot, aes(x = factor(year), y = fishing_hours / 1e3,
                       fill = gear_clean)) +
  geom_col(position = "stack", colour = "white", linewidth = 0.25) +
  scico::scale_fill_scico_d(palette = "batlow", name = "Gear type") +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.05))) +
  labs(
    title    = "Fishing Effort by Gear Type — Antarctic Peninsula (CCAMLR)",
    subtitle = glue("Annual apparent fishing hours (×10³), ",
                    "{min(HEATMAP_YEARS)}–{max(HEATMAP_YEARS)}"),
    x = NULL, y = "Fishing hours (thousands)",
    caption = GFW_CAPTION
  ) +
  theme_antarctic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot(gearplot)
```

---

```{r}
#| label: Save plots

ggsave(
  filename = "GFW_PeninsulaMap_panels.pdf",
  plot     = spatial_panels,
  width    = 12,
  height   = 8,
)
ggsave(
  filename = "GFW_AnnualMonthly_trend.pdf",
  plot     = temporal_compo,
  width    = 12,
  height   = 8,
)
ggsave(
  filename = "GFW_FisheriesGear.pdf",
  plot     = gearplot,
  width    = 12,
  height   = 8,
)

```


# Session information

```{r}
#| label: session-info
#| code-fold: false

sessionInfo()
```

::: {.callout-tip}
## Reproducibility & troubleshooting

**Cache files** in the working directory:

- `cache_heatmap_data.rds` — annual gridded effort (heatmaps + gear plot)
- `cache_trend_data.rds`   — monthly effort time series (trend section)

The notebook validates the cache on load and **automatically deletes and
re-fetches** if the file is empty or corrupt. To force a full refresh,
delete both `.rds` files before rendering.

**If fetches keep failing:**

1. Run `print_request = TRUE` in a manual `gfw_ais_fishing_hours()` call
   to see the exact URL being sent to the API.
2. Test the request interactively in R and check for HTTP 524 (timeout) or
   HTTP 429 (rate limit). For 429, wait ~90 seconds and re-render.
3. Contact GFW support at apis@globalfishingwatch.org with the printed
   request URL.
:::
